---
title: "Design of Experiments Coursework-34273638"
format: html
editor: visual
---

## Question 2

```{r}
#Setting up enviroment
library(tidyverse)
exercise <- data.frame(
 Subject = rep(1:3, rep(4, 3)),
 Day = rep(1:4, 3),
 Treatment = c(1, 4, 3, 2, 2, 1, 4, 3, 3, 2, 1, 4),
 Y = c(45, 27, 27, 15, 18, 32, 23, 15, 23, 16, 28, 20)
)

exercise <- exercise%>%
  mutate(Subject = factor(Subject,levels = c(1:3),labels=c(1:3)),
         Day = factor(Day, levels=c(1:4), labels=c(1:4)),
         Treatment = factor(Treatment, levels=c(1:4),labels=c(1:4)))
```

### Part a

Investigating differences between treatments and blocking factors Day and Subject:

```{r}
#from 3.7 Q2
aggregate(Y~Treatment,data=exercise,
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
aggregate(Y~Subject,data=exercise,
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
aggregate(Y~Day,data=exercise,
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
#aggregate(Y~Subject,data=exercise,FUN = mean)
#aggregate(Y~Day,data=exercise,FUN = mean)

#plots
boxplot(Y ~ Treatment,data=exercise)
boxplot(Y ~ Subject,data=exercise)
boxplot(Y ~ Day,data=exercise)


#with(exercise,
#  interaction.plot(Treatment, Day, Subject, Y, xlab = 'Treatment', ylab = 'Yield', trace.label = 'Block')
#)
```

From the boxplots, treatment 1 appears to result in a much larger average response compared to the other 3 treatments. However, the design is not a randomised complete block design (RCBD) as each treatment is not replicated in each block, thus we do not know if these differences are due to block effects instead. This is plausible as the day blocking effect varies substantially between days, with day 4 having a much smaller average response compared to day 1,2 or 3.

The unit block treatment model with 2 blocking factors is defined below:

$y_{ijk} = \mu + \alpha_{i} + \beta_{j} + \tau_{k} + \epsilon_{ijk}, \; \: \: i,k=\{1,2,3,4\}, \; \; j=\{1,2,3\}$

where $y_{ijk}$ is the time taken in seconds for 50 heartbeats of subject $j$ exercising on day $i$ using treatment $k$, $\mu$ is a constant parameter, $\alpha_{i}$ is the block effect of day $i$, $\beta_{j}$ is the block effect of subject $j$ and $\tau_{k}$ is the effect of treatment $k$. The errors follow a normal distribution $\epsilon_{ijk} \sim N(0,\sigma^{2})$ with mean 0 and constant variance, and are assumed independent for different experimental units. I Note that not all $y_{ijk}$ exist due to incompleteness of the design.

```{r}
#creating model
exercise.lm <- lm(Y ~ Day+Subject+Treatment,data=exercise)
anova(exercise.lm)
```

Fitting this model and performing an ANOVA test gives the above result. The treatment line of the ANOVA test compares models with and without the effects of treatment (but both models include blocking factors day and subject) by testing $H_{0}: \tau_{1}=\tau_{2}=\tau_{3}=\tau_{4}=0$ vs $H_{1}:\tau_{k} \ne 0$ for at least 1 $k \in \{1,2,3,4\}$. We get a significant p-value of 0.03012, thus we reject $H_{0}$ that there is no treatment difference at the 5% level indicating there is a significant effect of Treatment.

```{r}
#Investigating pairwise differences
exercise.emm <- emmeans::emmeans(exercise.lm, ~ Treatment)
pairs(exercise.emm)
```

As there is a significant effect of treatment we now estimate all pairwise differences using an experiment-wise error rate of 5%. Only the comparison between treatment 1 and 2 is significant using Tukey to adjust for multiple comparisons.

```{r}
#residual plots
standres <- rstandard(exercise.lm)
fitted <- fitted(exercise.lm)
par(mfrow = c(1, 2), pty = "s")
with(exercise, {
  plot(Subject, standres, xlab = "Subject", ylab = "Standarised residuals")
  plot(Day, standres, xlab = "Day", ylab = "Standarised residuals")
 par(mfrow = c(1, 2), pty = "s") 
  plot(fitted, standres, xlab = "Fitted value", ylab = "Standarised residuals")
})

par(pty = "s")
qqnorm(standres, main = "")

```

The residual plots show no large residual values with absolute values greater than 2 (which would indicate poor fit of the model). There is some evidence of unequal variation (in particular subjects' 2 variation is much lower than that of subjects 1 and 3). However, there is no obvious patterns with respect to the fitted values against the standardised residuals with a random scattering of points in the index plot of the residuals. Finally, the points of the qq-plot lie on a straight line which suggested the normality assumption is valid.

### Part b

```{r}
#Adding variables of two level factors
exercise <- exercise%>%
  mutate(Duration=ifelse(Treatment==1,1,ifelse(Treatment==2,1,3)),
         Speed=ifelse(Treatment==1,40,ifelse(Treatment==3,40,60)),
         Pedal=ifelse(Treatment==1,"hand",ifelse(Treatment==4,"hand","foot")),
         Duration= factor(Duration,levels=c(1,3),labels=c(1,3)),
         Speed=factor(Speed,levels=c(40,60),labels=c(40,60)),
         Pedal=factor(Pedal,levels=c("hand","foot"),labels=c("hand","foot")))

#Changing levels to high=+1 or low=-1
exercise_contrast <- exercise%>%
  mutate(Duration = ifelse(Duration==1,-1,1),
         Speed = ifelse(Speed==40,-1,1),
         Pedal = ifelse(Pedal=="hand",-1,1),
         Duration= factor(Duration,levels=c(-1,1),labels=c(-1,1)),
         Speed=factor(Speed,levels=c(-1,1),labels=c(-1,1)),
         Pedal=factor(Pedal,levels=c(-1,1),labels=c(-1,1)))
```

Adding the new factorial treatment effects gives the updated regression model below,

$y_{ijk_{1}k_{2}k_{3}} = \mu + \alpha_{i} + \beta_{j} + \tau_{k_{1}}^{S} + \tau_{k_{2}}^{D} + \tau_{k_{3}}^{P} + \epsilon_{ijk_{1}k_{2}k_{3}}, \; \: \: k_{w}=\{L,H\}, \;\;\; w = \{1,2,3\} \; \; j=\{1,2,3\}$

where $y_{ijk_{1}k_{2}k_{3}}$ is the time taken in seconds for 50 heartbeats of subject $j$ exercising on day $i$ using a combination of speed set to $k_{1}$, duration set to $k_{2}$ and pedal set to $k_{3}$. $\mu$ is a constant parameter, $\alpha_{i}$ is the block effect of day $i$, $\beta_{j}$ is the block effect of subject $j$, $\tau^{S}_{k_{1}}$ is the speed, $\tau^{D}_{k_{2}}$ is the duration and $\tau^{P}_{k_{3}}$ is the pedal with each set to low or high defined by the index. The errors follow a normal distribution $\epsilon_{ijk_{1}k_{2}k_{3}} \sim N(0,\sigma^{2})$ with mean 0 and constant variance, and are assumed independent for different experimental units. In the model, Duration=1, Speed=40 and Pedal=hand are defined as the low level of the two-level factors and Duration=3, Speed=60 and Pedal=foot as the high level.

```{r}
fact.lm <- lm(Y ~ Subject + Day + Duration +Speed + Pedal, data = exercise_contrast)
anova(fact.lm)
```

Investigating just the main effects after controlling for block effects, performing an ANOVA test after fitting the model above found both Speed and Pedal factorial effects to be significantly different from 0 with p-values of 0.03738 and 0.01376 respectively.

This makes intuitive sense as in part (a) we found that the only significant pairwise difference was between treatment 1 and 2 which varies by both speed and pedal factorial level (Treatment 1 has both set to low, meanwhile Treatment 2 has them both set to high). But does vary by duration (both treatments have duration set to low). Using principle of effect hierarchy thus main effects are more likely to have large effect than 2 factor interactions we can say the significant effect of treatment 1 compared to treatment 2 was due to factorial effects of speed and pedal changing and therefore they must also have a significant effect as was found. Furthermore, as no treatment comparsions in which duration changed level was significant, duration does not have a large and thus significant effect on the response (as seen by non-significant p-value).

```{r}
2*coef(fact.lm)[7:9]
```

The factorial effects are estimated to be -14.25, -20.75 and -3 for changing speed, pedal and duration respectively from low to high.

### Part c

Due to the design, the contrast used to estimate the two-factor interaction is exactly the same as the main effect of the factor not included in the interaction.

This is due to only 4 treatments combinations being investigated in the deisgn, a full factorial design would have $2^{3}=8$ different treatments. Thus the design is a fractional factorial design with certain factorial effects being aliased. In particular, the aliasing scheme for main and two-factor interactions is S=PD, P=SD and D=DS for S the speed, P the pedal and D the duration. Therefore, to estimate the effect of the two-factor interaction we must assume the main effect of the third factor is zero (as we are actually estimating the linear combination of the aliased main effect and the two-factor interaction). I note this goes against the principle of effect hierarchy thus the estimates obtained are more likely to be biased due to the assumption that the main effect is equal to 0 being incorrect.

With the factorial effects have been aliased with S=PD, P=SD and D=DS for S the speed, P the pedal and D the duration. Therefore, to estimate the effect of the two-factor interaction we must assume the main effect of the third factor is zero (as we are actually estimating the linear combination of the aliased main effect and the two-factor interaction). I note this goes against the principle of effect hierarchy thus the estimates obtained are more likely to be biased due to the main effect equal to 0 assumption being incorrect.

```{r}
fact.npedal.lm <- lm(Y ~Subject + Day + Duration+Speed+Duration:Speed, data = exercise_contrast)

2*coef(fact.npedal.lm)[9]

fact.nspeed.lm <- lm(Y ~Subject + Day + Duration+Pedal+Duration:Pedal, data = exercise_contrast)
2*coef(fact.nspeed.lm)[9]

fact.nduration.lm <- lm(Y ~Subject + Day + Pedal+Speed+Pedal:Speed, data = exercise_contrast)
2*coef(fact.nduration.lm)[9]
```

To estimate the two-factor interactions I exclude the third factor from the linear model and replace it with the two-factor interaction. The estimates of the two-factor interactions when both factors set to high are D:S = 41.5, D:P = 28.5 and P:S = 6 but these results should be interpreted with some skepticism due to the issues mentioned earlier. Furthermore, these results cannot be considered together as each model implies the other 2 are incorrect as a factor included in that model should be zero from the assumptions of the model. From part (b) the most likely model includes speed, ,pedal and their interaction as these were the two main effects that were significant (and thus non-zero) and due to principle of effect heredity their interaction is more likely to be important.

## Older stuff (Check through this against current part b/c)

Note that this model is not saturated as the each treatment that is observed occurs multiple times when not considering the blocking factors??

This comparison may not be 'fair' as when estimating the main effects the combination of the other factors are different for the high and low levels of the factor interested in. Although each other factor occurs a equal number of times in its low and high category for each treatment being investigated.???

\*Model uses 4 df to estimate the 3 factors and the intercept, then 3 for the 2 factor interactions, then for blocks uses 2??? (check this in chapter 5/6).

```{r}
## Extra stuff :P

# #testing against models without factor (same as the anova test above but just doing explictly)
# 
# fact.lm.1 <- lm(Y ~ Subject + Day + Speed + Pedal, data = exercise_contrast)
# fact.lm.2 <- lm(Y ~ Subject + Day + Duration + Pedal, data = exercise_contrast)
# fact.lm.3 <- lm(Y ~ Subject + Day + Duration + Speed, data = exercise_contrast)
# 
# anova(fact.lm, fact.lm.1)
# anova(fact.lm, fact.lm.2)
# anova(fact.lm, fact.lm.3)

#following code from section 5
X <- model.matrix( ~ Subject + Day + (Duration+Speed+Pedal)^3, data = exercise_contrast)
#note that there are 12 different blocks (3*4=12) and each one occurs once

#Defined hand as the low level, speed=40 as low level and duration=1 as low level
#all low levels of factor coded as -1 and all high levels as +1
main_effect_contrasts <- matrix(c(-1,-1,-1,
                                  1,1,-1,
                                  1,-1,1,
                                  -1,1,1,
                                  -1,1,1,
                                  -1,-1,-1,
                                  1,1,-1,
                                  1,-1,1,
                                  1,-1,1,
                                  -1,1,1,
                                  -1,-1,-1,
                                  1,1,-1)/6,c(12,3),byrow=T)
#Each column is the main effect contrast for duration, speed and pedal respectively

t(as.matrix(main_effect_contrasts)) %*% exercise$Y

#Averaging the main effects over the blocking factors. The response decreases at the higher duration of 3 minutes by 3.167 compared to the lower duration of 1 minute. The response decreases at the higher speed of 60 rpm by 8.5 compared to the lower speed of 40 rpm. The response decreases when using the foot pedal by 10.167 compared to using the hand pedals.
```
