---
title: "Untitled"
format: html
editor: visual
---

## Part 2

```{r}

library(tidyverse)
exercise <- data.frame(
 Subject = rep(1:3, rep(4, 3)),
 Day = rep(1:4, 3),
 Treatment = c(1, 4, 3, 2, 2, 1, 4, 3, 3, 2, 1, 4),
 Y = c(45, 27, 27, 15, 18, 32, 23, 15, 23, 16, 28, 20)
)

exercise <- exercise%>%
  mutate(Subject = factor(Subject,levels = c(1:3),labels=c(1:3)),
         Day = factor(Day, levels=c(1:4), labels=c(1:4)),
         Treatment = factor(Treatment, levels=c(1:4),labels=c(1:4)))

head(exercise)
```

Investigating differences:

```{r}
aggregate(Y~Treatment,data=exercise,FUN = mean)
aggregate(Y~Subject,data=exercise,FUN = mean)
aggregate(Y~Day,data=exercise,FUN = mean)
```

$y_{ijkl} = \mu + \alpha_{i} + \beta_{j} + \tau_{k} + \epsilon_{ijkl}$

Here $\alpha$ represents the blocking factor day, $\beta$ representants the blocking factor subject and $\tau$ represents the treatment effect.

```{r}
#creating model
exercise.lm <- lm(Y ~ Day+Subject+Treatment,data=exercise)
anova(exercise.lm)
```

Estimating all pairwise differences.

```{r}
#Investigating pairwise differences
exercise.emm <- emmeans::emmeans(exercise.lm, ~ Treatment)
pairs(exercise.emm)
```

only the coparison between treatment 1 and 2 is significant at the 5% level with a p-value of 0.0252 after adjustment of multiple comparisons using Tukey.

```{r}
#redsidual plots
plot()
```
